# Linear Models

```{r, include = FALSE}
# For next time
# - DO NOT interpret non-significant results
# - Add significance to interpretations
# - Add explanations of ANOVA levels

knitr::opts_chunk$set(dpi = 150, fig.width = 8, out.width = "90%",
                      fig.align = "center", fig.asp = 0.45, cache = FALSE, 
                      warning = FALSE)

library(knitr)
hook_output <- knit_hooks$get("output")
knit_hooks$set(output = function(x, options) {
   lines <- options$output.lines
   if (is.null(lines)) {
     return(hook_output(x, options))  # pass to default hook
   }
   x <- unlist(strsplit(x, "\n"))
   x <- x[lines]

   # paste these lines together
   x <- paste(c(x, ""), collapse = "\n")
   hook_output(x, options)
 })

library(here)
library(flair)
library(glue)
library(ggplot2)
library(palmerpenguins)
library(knitr)
library(patchwork)
options(width = 100)


colourize <- function(x, colour) {
  if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'>%s</span>", colour,
      x)
  } else x
}

purple <- "#440154"
teal <- "#277F8E"
```

## Preamble

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
library(patchwork) # putting figures together
library(car) # For Anova() and vif() functions
```

Here we're talking about *linear* relationships. There are plenty of other types of models (e.g., exponential) which deal with non-linear relationships or a mix of linear and non-linear (e.g., GAMs).


> **Disclaimer:**
>
> The models included here are examples, they're not the best (or most sensible) questions to ask, and often explore correlated features rather than features that we might actually assume causation.
>
> Technically these might be better explored using Standardised Major Axis Estimate (SMA) in the smatr package. But for now, the'll suffice `r emo::ji("grin")`
>

## General Workflow

1. Decide on your question(s)
2. Preliminary data exploration
3. Decide on a model (linear model, generalized linear model, mixed model, etc.)
4. Run model
5. Check model diagnostics
6. Interpret model


## (Simple) LMs in R

### Requirements {-}
- **Continuous response** (`y`)
  - no count data, no binary (binomial) data, no proprotions (although could be transformed)
- **Continuous or categorical predictors** (`x1`, `x2`, etc.)

### Assumptions {-}
- **Linear relationship** between `y` and `x`s.
- **Independence** - Observations are independent (i.e. residuals are independent)
  - no hierarchical sampling, no nested designs, variables are based on each other
- **Normality** - Residuals are normally distributed
- **Constant variance** - There is constant variance in residuals (no heteroscedasticity)

### Cautions {-}

- **Multicollinearity**
- **Influence**

> See [Diagnostics](appendix-model-diagnostics.html) section for more details

### Types of models {-}

- **Regression** - Continuous predictor (`x`)
- **Multiple regression** - More than one continuous predictor 
  (`x1`, `x2`, ...)
- **ANOVA** - Categorical predictor(s) (`x1`, `x2`, ...)
- **ANCOVA** - Mix of categorical and continuous predictors (`x1`, `x2`, ...)


### Run models with `lm()` {-}

:::{.align-centre .large}
`lm(y ~ x1 + x2, data = data)`
:::

> R will figure out the underlying model type (regression, ANOVA, ANCOVA etc.)



Store your model as an object, here `m`

```{r}
m <- lm(body_mass_g ~ bill_depth_mm, data = penguins)
```


`m` is a model object
```{r}
class(m)
```

This contains all the information about the model

Use `summary()` to show summary table:

```{r}
summary(m)
```

Use `anova()` to show Anova table (Type I), or, better, `Anova()` (from the car package) to use type II or III:
```{r}
Anova(m, type = "III")
```

### Summary tables vs. ANOVA tables {-}

> **NOTE**  
> Even though traditionally a summary table is used to evaluate a regression and 
> an ANOVA table is used to evaluate an ANOVA or an ANCOVA, they can both be used
> to assess any linear model, they just ask slightly different questions


**Summary tables**

Summary tables show the results of very specific statistical questions.
The precise interpretation of these results depends on whether the parameters 
are continuous, categorical, or a mix, and whether there are interactions or not.
(see specific sections for interpretations).


**ANOVA tables**

Broadly, an ANOVA table asks whether there is an **overall** effect of any 
parameter, but it **does not** specify what that effect is.

This means that for *continuous* parameters you don't know if the effect is 
positive or negative. 

For *categorical* parameters you don't know which category has what effect
(but see Post-Hoc analyses which can be used to tease this apart). 

## Regressions

- Continuous response (`y`) and **continuous** predictor(s) (`x`)
- What might we be interested in?

```{r, echo = FALSE, fig.asp = 0.8, fig.width = 7, message = FALSE}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  stat_smooth(method = "lm") +
  ggtitle("Is penguin body mass a function of skeletal size?")

ggplot(data = penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  stat_smooth(method = "lm") +
  ggtitle("Can body mass be predicted by flipper length and bill length?")
```


### Create Model {-}

To see if these relationships are significant we run/create a linear model...

```{r}
m <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins)
```

> **Wait!** We shouldn't interpret until we've checked our diagnostics


### Model Diagnostics {-}

- Checking Assumptions
  - Normality (of residuals)
  - Constant Variance (no heteroscedasticity)
- Other cautions
  - Influential observations (Cook's D)
  - Multiple collinearity (with more than one `x` or explanatory variables)


> See [Diagnostics](appendix-model-diagnostics.html) section for more details


```{r}
d <- data.frame(residuals = residuals(m),    
                std_residuals = rstudent(m), 
                fitted = fitted(m),          
                cooks = cooks.distance(m)) %>%
  mutate(obs = 1:n())
```


#### Normality and Heteroscedasticity {-}


```{r, fig.asp = 0.7, fig.width = 8}
g1 <- ggplot(data = d, aes(x = std_residuals)) +
  geom_histogram(bins = 20) +
  ggtitle("Histogram of residuals")

g2 <- ggplot(data = d, aes(sample = std_residuals)) +
  stat_qq() +
  stat_qq_line() +
  ggtitle("QQ Normality plot of residuals")

g3 <- ggplot(d, aes(x = fitted, y = std_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0) +
  ggtitle("Heteroscedacity: Residuals vs. fitted")

(g1 + g2) / g3
```


#### Influence {-}

**Cook's D** 
```{r}
ggplot(d, aes(x = obs, y = cooks)) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_hline(yintercept = 4/nrow(penguins), 
             linetype = "dashed") +
  ggtitle("Cook's D: Influential observations")
```


> Pretty good!



#### Multicollinearity (collinearity) {-}

Look at our two explanatory variables
```{r, warning = FALSE, message = FALSE, fig.asp = 0.7, fig.width = 6}
ggplot(data = penguins, aes(x = flipper_length_mm, y = bill_length_mm)) +
  geom_point() +
  stat_smooth(method = "lm")
```

> Definitely correlated.
> We need to investigate if it interferes with the model.



Use `vif()` function from `car` package^[vif = variance inflation factor; Can be interpreted as how much influence the variable has on the model]

```{r, message = FALSE}
library(car)
vif(m)
```

> Hmm, that's pretty good (looking for < 10).
> 
> So we can assume that the correlation is not a problem for how we interpret the model


### Summary Tables {-}

```{r}
m <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins)
summary(m)
```


```{r lm_summary2, echo = FALSE, eval = FALSE}
summary(m)
```

That's a lot of information! Let's break it down.

**Model**
```{r, ref.label = "lm_summary2", output.lines=2:4, echo = FALSE}
```

**Missing observations**
```{r, ref.label = "lm_summary2", output.lines=19:19, echo = FALSE}
```

**R<sup>2</sup> and adjusted R<sup>2</sup>**

- Adjusted for the number of parameters

```{r, ref.label = "lm_summary2", output.lines = 20, echo = FALSE}
```


**Effects**

```{r, ref.label = "lm_summary2", output.lines=11:14, echo = FALSE}
```

```{r, echo = FALSE}
tribble( ~Column,       ~Description,
        "`Estimate`",   "Intercept/Slope of the effect",
        "`Std. Error`", "Variability in the estimates",
        "`t value`",    "Test statistic^[Think of it as a holistic combination of estimate and variability]",
        "`Pr(>|t|)`",   "**P-value**, significance of the results^[Probability of getting `t-value` by chance]") %>%
  kable(format = "html", escape = FALSE)
```

### Interpreting effects {-}

```{r, ref.label = "lm_summary2", output.lines=11:14, echo = FALSE}
```

```{r, echo = FALSE, results = "asis"}
broom::tidy(summary(m)) %>%
  mutate(estimate = round(estimate, 2),
         p.value = if_else(p.value < 2.2e-16, "< 2.2e-16", as.character(round(p.value, 3)))) %>%
  glue_data(
    "- **Intercept**",
    "  - Significant (P = {p.value[1]})",
    "  - Penguins with a flipper length of 0 mm and a bill length of 0 mm are predicted to have a body mass of {estimate[1]} g", 
    
    "- **Flipper Length**",
    "  - Significant (P = {p.value[2]})",
    "  - For each 1 mm increase in flipper length, body mass is predicted to increase by {estimate[2]} g", 
    
    "- **Bill Length**",
    "  - Non-significant (P = {p.value[3]})",
    "  - So we don't interpret)",
    .sep = "\n")
```

(**Note**: 2e<sup>-16</sup> = 0.0000000000000002, R uses this as the smallest number)

> **Therefore**
> 
> There is a significant relationship between flipper length and body mass.  
> But *not* between bill length and body mass (when including and controlling for
> flipper length)

### Interpreting the effects as a model {-}

First we have the general equation for a line^[(remember highschool?)]. 

$y = mx + b$

- $m$ = slope of $x$ 
- $b$ = intercept    


For two explanatory variables

$y = m_1x_1 + m_2x_2 + b$

For our specific model

$y = `r round(coef(m)[2], 2)`x_1 + `r round(coef(m)[3], 2)`x_2 + (`r round(coef(m)[1], 2)`)$




### Extra {-}

> Why no effect of Bill Length? {-}


```{r, message = FALSE, echo = FALSE, fig.asp = 1, fig.width = 5, out.width = "100%"}
ggplot(data = penguins, aes(x = bill_length_mm, y = body_mass_g)) +
  geom_point() +
  stat_smooth(method = "lm") 
```

```{r, highlight.output = 12}
m <- lm(body_mass_g ~ bill_length_mm, data = penguins)
summary(m)
```

- Our hypothesis is of *causation* but really, this is just correlation
- Flipper length is the 'better' predictor of body mass
- When flipper length in the model, no extra variation explained by bill length
- When flipper length *not* in the model, some variation left to be explained

```{r, highlight.output = 12}
m <- lm(body_mass_g ~ bill_length_mm, data = penguins)
summary(m)
```


## ANOVAs 

- Continuous response (`y`) and categorical predictor(s) (`x`)
- What might we be interested in?

```{r, echo = FALSE, fig.asp = 0.8, fig.width = 7, message = FALSE}
ggplot(data = drop_na(penguins), aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  ggtitle("Are different species different sizes?")

ggplot(data = drop_na(penguins), aes(x = sex, y = body_mass_g)) +
  geom_boxplot() +
  ggtitle("Are male penguins larger than female penguins?")
```

### Create Model {-}

To see if these relationships are significant we run/create a linear model...

```{r}
m <- lm(body_mass_g ~ species + sex, data = penguins)
```

> But first diagnostics...


### Model Diagnostics {-}

> As with regressions...

- Checking Assumptions
  - Normality (of residuals)
  - Constant Variance (no heteroscedasticity)
- Other cautions
  - Influential observations (Cook's D)
  - Multiple collinearity (with more than one `x` or explanatory variables)


> See [Diagnostics](appendix-model-diagnostics.html) section for more details


```{r}
d <- data.frame(residuals = residuals(m),    
                std_residuals = rstudent(m), 
                fitted = fitted(m),          
                cooks = cooks.distance(m))   

d <- mutate(d, obs = 1:n())
```


```{r}
head(d)
```


#### Normality and Heteroscedasticity {-}


```{r, fig.asp = 0.7, fig.width = 8}

g1 <- ggplot(data = d, aes(x = std_residuals)) +
  geom_histogram(bins = 20)

g2 <- ggplot(data = d, aes(sample = std_residuals)) +
  stat_qq() +
  stat_qq_line()

g3 <- ggplot(d, aes(x = fitted, y = std_residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)

(g1 + g2) / g3
```

#### Influence {-}

**Cook's D**
```{r, fig.asp = 0.50, fig.width = 5}
ggplot(d, aes(x = obs, y = cooks)) +
  geom_point() +
  geom_hline(yintercept = 1, linetype = "dotted") + #<<
  geom_hline(yintercept = 4/nrow(penguins), 
             linetype = "dashed")
```


> Overall not too bad


#### Multicollinearity (collinearity) {-}

- Only relevant with **more than one explanatory variable**
- Sometimes explanatory variables are so correlated they interfere with the model
- Correlations between variables *might* be problematic (but not necessarily)

Use `vif()` function from `car` package^[vif = variance inflation factor; Can be interpreted as how much influence the variable has on the model]

```{r, message = FALSE}
library(car)
vif(m)
```

Here we consider the `GVIF^(1/2*Df))` value<sup>*</sup>

Looks good!

(\* See `?vif` and the reference therein: Fox, J. and Monette, G. (1992) Generalized collinearity diagnostics. JASA, 87, 178–183.)


### Summary Tables {-}

```{r}
m <- lm(body_mass_g ~ species + sex, data = penguins)
summary(m)
```


```{r anova_summary, echo = FALSE, eval = FALSE}
summary(m)
```

That's a lot of information! Let's break it down.

**Model**
```{r, ref.label = "anova_summary", output.lines=2:3, echo = FALSE}
```

**Missing observations**
```{r, ref.label = "anova_summary", output.lines=19:19, echo = FALSE}
```

**R<sup>2</sup> and adjusted R<sup>2</sup>**

- Adjusted for the number of parameters

```{r, ref.label = "anova_summary", output.lines = 20, echo = FALSE}
```


**Effects**

```{r, ref.label = "anova_summary", output.lines=10:14, echo = FALSE}
```

```{r, echo = FALSE}
tribble( ~Column,       ~Description,
        "`Estimate`",   "Intercept/Slope/Mean difference of the effect",
        "`Std. Error`", "Variability in the estimates",
        "`t value`",    "Test statistic^[Think of it as a holistic combination of estimate and variability]",
        "`Pr(>|t|)`",   "**P-value**, significance of the results^[Probability of getting `t-value` by chance]") %>%
  kable(format = "html", escape = FALSE)
```

### Interpreting effects {-}

```{r, ref.label = "anova_summary", output.lines=11:14, echo = FALSE}
```


- `Estimate`
    - Treatment contrasts
    - Average *differences* among categories compared to the base category
    
- `Std. Error` 
    - Variability in the estimates
    
- `t value`
   - Test statistic

- `Pr(>|t|)` 
    - **P-value**, significance of the *differences*
    - Probability of getting `t-value` by chance

> Easier to interpret estimates if we consider a simpler model


**A simpler model**

```{r}
m <- lm(body_mass_g ~ species, data = penguins)
```

```{r anova_summary2, eval = FALSE}
summary(m)
```


```{r, ref.label = "anova_summary2", echo = FALSE, output.lines = 11:13}
```


Effect of **Species**

- `(Intercept)` represents base category (i.e. Adelie penguins)
- Adelie have mean body mass of `r round(coef(m)[1], 2)` g 
- On average, Chinstrap penguins are `r round(coef(m)[2], 2)` g heavier than Adelie penguins
- On average, Gentoo penguins are `r round(coef(m)[3], 2)` g heavier than Adelie penguins


> Back to original model

```{r}
m <- lm(body_mass_g ~ species + sex, data = penguins)
```


```{r, ref.label = "anova_summary", echo = FALSE, output.lines = 11:14}
```

Effect of **Species** and **Sex**

- `(Intercept)` represents base category but is a combination of factors
- Much more complicated to interpret
- Comparisons are often not of interest anyway 
  (unless you've set up contrasts, which are advanced stats but awesome!)

> So let's look at ANOVA tables instead

### ANOVA Tables {-}

#### Type I

```{r anova}
anova(m)
```

#### Type III {-}
```{r anova3}
Anova(m, type = "3")
```

`Anova()` is from the car package. 


Overall effects of **Species** and **Sex**

- Yes there are differences among **Species** (P < 2.2e<sup>-16</sup>)
- Yes there are differences between **Sexes** (P < 2.2e<sup>-16</sup>)

> Not a whole lot of information... <br>We'll have to do some <strong>[Post-Hoc](#post-hoc-tests)</strong> tests!


#### Type I vs. III {-}

**Why one or the other?**

```{r, highlight.output = 5:6}
m1 <- lm(body_mass_g ~ species + sex, data = penguins) #<<
anova(m1)
```

```{r, highlight.output = 5:6}
m2 <- lm(body_mass_g ~ sex + species, data = penguins)#<<
anova(m2)
```

- For Type I ANOVAs, order matters with unbalanced samples
    - See that `Sum sq`, `Mean Sq` and `F value` all differ between the models
- Here, pretty minor, but important to remember with greater unbalances



```{r, ref.label = "anova3", echo = FALSE}
```

- Type III ANOVAs are not dependent on variable order


> Type II ANOVAs do exist as well, but generally we use Type III in natural sciences


### Post-Hoc tests {-}

When you would like to understand differences between 2+ levels in a categorical variable, you'll need to do post-hoc tests.

Note:

- We do post-hoc tests *after* having found a *significant* of the variable in the ANOVA table.
- There's no point in doing post-hoc tests unless you have a variable with more than three levels, OR you're interested in understanding the differences among categorical variables that also have an interaction
- In our example, we don't need to do post-hoc tests for sex, because there are only two levels, `male` and `female`. If there is a difference, then males are different than females, no question.




