# Appendix Model Diagnostics {-}

```{r, message = FALSE}
library(tidyverse)
library(palmerpenguins)
library(car)

library(lme4)
library(glmmTMB)
```

There are many different kinds of model checking/diagnostics that need to be used for different kinds of models.

a) normally distributed residuals (gaussian)
b) constant variance in residuals (gaussian)
c) checking for influential observations (all)
d) assessing multiple collinearity (all)
e) overdispersion (poisson and binomial)
f) DHARMa package to assessing model fit using simulated residuals (all)

> In this section we will cover how to run and interpret various diagnostics
> that can be used for different types of models.

## Diagnostics by hand {-}

There are many functions for checking diagnostics in R but generally they only work with specific model types.

However, many diagnostics simply require **fitted values** (model predictions)
and **residuals** (model errors) which we can extract and use to create our own 
diagnostic plots.

### Extracting parameters {-}

```{r}
m <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm, data = penguins)
```


- **`residuals()`**
- **`fitted()` values**
- **`cooks.distance()`**
- **`as.numeric(names(residuals()))`** Observation number 
  - (accounting for any omitted, missing values)

```{r}
d <- tibble(residuals = residuals(m),    
            fitted = fitted(m),          
            cooks = cooks.distance(m),
            obs = as.numeric(names(residuals)))
```

```{r}
head(d)
```

## Normality {-}

QQ Normality plot of residuals
```{r, fig.asp = 0.65, fig.width = 4}
ggplot(data = d, aes(sample = residuals)) +
  stat_qq() +
  stat_qq_line()
```

## Variance {-}

Check heteroscedasticity

```{r, fig.asp = 0.7, fig.width = 4}
ggplot(d, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)
```


## Influence (Cook's D) {-}

- When an observation has an unusually high effect on the model
- If you omit an observation and the entire model changes, that observation is said to have a lot of *influence*
- Cook's D is a way of measuring influence



```{r, fig.asp = 0.5, fig.width = 5}
ggplot(d, aes(x = obs, y = cooks)) +
  geom_point()
```

Use ggiraph to get interactive tooltips (hover over a point to find which it is!)
```{r, fig.asp = 0.5, fig.width = 5}
library(ggiraph)
g <- ggplot(d, aes(x = obs, y = cooks)) +
  geom_point_interactive(aes(tooltip = obs))
girafe(ggobj = g, width_svg = 6, height_svg = 3)
```

### Mixed models {-}

In addition to the observation level effects (above) you'll also want to take 
look at the random effects

```{r}
m <- lmer(body_mass_g ~ flipper_length_mm + bill_length_mm + (1|island), data = penguins)

r <- as.data.frame(ranef(m))
```

QQ Normality plot of random effects
```{r}
ggplot(data = r, aes(sample = condval)) +
  stat_qq() +
  stat_qq_line()
```


## Multicollinearity {-}

- Only relevant with **more than one explanatory variable**
- Sometimes explanatory variables are so correlated they interfere with how you might interpret them
  - Note that this means that if variables are only included as covariates 
  (i.e. to control for their effects, not because you want to interpret them), 
  then they can be highly correlated because their correlation won't affect other
  parameters in the model^[(ref:collinearity)]
- Correlations between variables *might* be problematic (but not necessarily)

Use `vif()` function from `car` package^[vif = variance inflation factor; Can be interpreted as how much influence the variable has on the model]

```{r, message = FALSE}
vif(m)
```


> Interpreting VIF
>  
> - 1 means no correlation
> - 1-5 means moderate but probably not a problem
> - 5-10 is problematic, p-values may no longer be valid
> - >10 FIX!


## Overdispersion {-}

> Coming!



## DHARMa {-}

```{r}
library(DHARMa)

r <- simulateResiduals(m, plot = TRUE)
```

```{r}
m <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + species, data = penguins)
r <- simulateResiduals(m, plot = TRUE)

m <- lm(body_mass_g ~ flipper_length_mm + bill_length_mm + species * sex, data = penguins)
r <- simulateResiduals(m, plot = TRUE)
```

### DHARMa is applicable to most linear models {-}

```{r}
penguins <- mutate(penguins, 
                   size = flipper_length_mm > median(flipper_length_mm, na.rm = TRUE))

m <- glm(size ~ species, family = "binomial", data = penguins)
r <- simulateResiduals(m, plot = TRUE)
```
